---
title: "EPID 674: Data Management in R"
subtitle: "Homework 4 Template"
author: "Insert your name here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r load_packages, echo=T, results='hide', message=F, warning=F}

##### Load packages for this assignment, do this every time
library(tidyverse)
library(here)
library(ggpubr)
library(viridis)
library(RColorBrewer)
```

# Load data
```{r read_data}

# Load the saved R data
load(here("nhanes_homework.rda"))
```



# 1. Create a bar plot 

### Create a bar plot where cognitive impairment status (variable: MCI) is on the x-axis and the bar fill is by race/ethnicity (variable: race_eth). Try updating the bar colors. 

```{r barplot, warning = F}

# Insert code for problem 1 here

```

Interpret the figure: 


# 2. Create a histogram or density plot

### Make a histogram or density plot with the digit symbol substitution test score (variable: CFDDS) on the x-axis. Have the fill and/or color of the plot by category of educational attainment.

```{r histogram_density_plot, warning = F}

# Insert code for problem 2 here

```

Interpret the figure: 



# 3. Create a boxplot or violin plot

### Make a boxplot or violin plot with the digit symbol substitution test score (variable: CFDDS) on the y-axis. Split the boxes or violins by the age group (variable: age_group) on the x-axis. Try filling the boxes with the color of your choice

```{r boxplot, warning = F}

# Insert code for problem 3 here

```

Interpret the figure: 



# 4. Create a scatterplot with a line of best fit

### Create a scatter plot with poverty-income ratio (variable: INDFMPIR) on the x-axis and digit symbol substitution test (variable: CFDDS) on the y-axis. Add a line of best fit for this relationship.

```{r scatterplot, warning = F}

# Insert code for problem 4 here

```

Interpret the figure: 


# 5. Create a multipaneled plot

### Create a multipaneled plot containing each of the previous plots from questions 1-4 (a barplot, a histogram, a boxplot or violin plot, and a scatterplot)

```{r multiplaneled, warning = F}

# Insert code for problem 5 here

```

Interpret the figure: 
