---
title: "EPID 674: Data Management in R"
subtitle: "Homework 4 Answer Key"
author: "Kelly Bakulski"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r load_packages, echo=T, results='hide', message=F, warning=F}

##### Load packages for this assignment, do this every time
library(tidyverse)
library(here)
library(ggpubr)
library(viridis)
library(RColorBrewer)
```

# Load data
```{r read_data}

# Load the saved R data
load(here("nhanes_homework.rda"))
```



# 1. Create a bar plot 

### Create a bar plot where cognitive impairment status (variable: MCI) is on the x-axis and the bar fill is by race/ethnicity (variable: race_eth). Try updating the bar colors. 

```{r barplot, warning = F}

# Optional, drop the participants with missing cognition values
nhanes_cognition <- nhanes_homework %>%
  drop_na(MCI)

# Make the bar plot
ggplot(nhanes_cognition,
       aes(x = MCI,
           fill = race_eth)) +
  geom_bar() +
  labs(title = "Barplot by race/ethnicity and cognitive status",
       x = "Cognitive impairment status",
       y = "Number of Participants",
       fill = "Race/ethnicity of participant") + 
  scale_fill_brewer() 

```

Interpret the figure: There appears to be a higher proportion of Non-Hispanic White participants in the no impairment category relative to the mild cognitive impairment category.


# 2. Create a histogram or density plot

### Make a histogram or density plot with the digit symbol substitution test score (variable: CFDDS) on the x-axis. Have the fill and/or color of the plot by category of educational attainment.

```{r histogram_density_plot, warning = F}

# Optional, drop participants missing education
nhanes_education <- nhanes_homework %>%
  drop_na(education)

# Overlapping histograms
ggplot(nhanes_education,
       aes(x = CFDDS,
           fill = education,
           y = ..density..)) + 
  geom_histogram(binwidth = 5,
                 position = "identity", #default position is stack 
                 alpha = 0.2)  +
  geom_density(aes(color = education),
               alpha = 0) +
  labs(x = "Digit Symbol Substitution Test Score",
       y = "Proportion of participants",
       title = "Histogram of cognition score by educational attainment",
       fill = "Educational attainment",
       color = "Educational attainment")

```

Interpret the figure: Within each educational attainment group, the distribution of digit symbol substitution score is relatively normally distributed. 



# 3. Create a boxplot or violin plot

### Make a boxplot or violin plot with the digit symbol substitution test score (variable: CFDDS) on the y-axis. Split the boxes or violins by the age group (variable: age_group) on the x-axis. Try filling the boxes with the color of your choice

```{r boxplot, warning = F}

# Boxplot split by age groups
ggplot(nhanes_homework,
       aes(x = age_groups,
           y = CFDDS)) +
  geom_boxplot(fill = "dodgerblue") +
  labs(x = "Age group",
       y= "Digit Symbol Substitution Test score",
       title = "Boxplot of cognition score by age group")


```

Interpret the figure: Yes, there does appear to be a relationship between age group and digit symbol substitution test. Higher age gropus have lower digit symbol substitution test scores. 



# 4. Create a scatterplot with a line of best fit

### Create a scatter plot with poverty-income ratio (variable: INDFMPIR) on the x-axis and digit symbol substitution test (variable: CFDDS) on the y-axis. Add a line of best fit for this relationship.

```{r scatterplot, warning = F}

# Add a linear fit
ggplot(nhanes_homework,
       aes(x = INDFMPIR,
           y = CFDDS)) +
  geom_point(alpha = 0.5,
             color = "seagreen") +
  geom_smooth(method = lm) +
  labs(x = "Poverty to income ratio",
       y = "Digit Symbol Substition Test score",
       title = "Scatterplot of poverty income ratio by digit symbol substitution test")

```

Interpret the figure: Yes, there does appear to be a relationship between poverty to income ratio and digit symbol substitution test score. Higher values of poverty to income ratio have  higher digit symbol substitution test scores.


# 5. Create a multipaneled plot

### Create a multipaneled plot containing each of the previous plots from questions 1-4 (a barplot, a histogram, a boxplot or violin plot, and a scatterplot)

```{r multiplaneled, warning = F}

bar <- ggplot(nhanes_cognition,
       aes(x = MCI,
           fill = race_eth)) +
  geom_bar() +
  labs(title = "Barplot by race/ethnicity and cognitive status",
       x = "Cognitive impairment status",
       y = "Number of Participants",
       fill = "Race/ethnicity of participant") + 
  scale_fill_brewer() 

hist <- ggplot(nhanes_education,
       aes(x = CFDDS,
           fill = education,
           y = ..density..)) + 
  geom_histogram(binwidth = 5,
                 position = "identity", #default position is stack 
                 alpha = 0.2)  +
  geom_density(aes(color = education),
               alpha = 0) +
  labs(x = "Digit Symbol Substitution Test Score",
       y = "Proportion of participants",
       title = "Histogram of cognition score by educational attainment",
       fill = "Educational attainment",
       color = "Educational attainment")

box <- ggplot(nhanes_homework,
       aes(x = age_groups,
           y = CFDDS)) +
  geom_boxplot(fill = "dodgerblue") +
  labs(x = "Age group",
       y= "Digit Symbol Substitution Test score",
       title = "Boxplot of cognition score by age group")


scatter <- ggplot(nhanes_homework,
       aes(x = INDFMPIR,
           y = CFDDS)) +
  geom_point(alpha = 0.5,
             color = "seagreen") +
  geom_smooth(method = lm) +
  labs(x = "Poverty to income ratio",
       y = "Digit Symbol Substition Test score",
       title = "Scatterplot of poverty income ratio by digit symbol substitution test")


ggarrange(bar, hist, box, scatter, 
          nrow = 2, 
          ncol = 2)

```

Interpret the figure: The histogram of cognition score by educational attainment is in the upper right hand corner.
