---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Homework 5A"
author: "Insert Your Name Here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(foreign)
library(epiDisplay)
library(stats)
library(Hmisc)
library(gmodels)

directory <- "/cloud/project/"
```

# Using the dataset “bpa.sas7bdat”, answer the following. Include appropriate titles, labels and legends in all graphs.

# T2DM can be defined as hemoglobin A1c ≥ 6.5% OR use of diabetes medication. Note: participants with high A1c and taking medication should also be considered T2DM cases.

# Example code to create the T2MD variable:
# bpa$t2dm<-ifelse(bpa$a1c>=6.5|bpa$dmmed==1, 1, 0)


# Problem 1
## Create a bar plot of T2DM by gender AND a barplot of T2DM by race/ethnicity. Try different bar colors. 

```{r problem 1}
#Insert code for problem 1 here
```
# Include plot for problem 1 here

# Problem 2
## Make a histogram for body mass index (bmi). What is the distribution of BMI? Does the distribution of log-transformed BMI look closer to the normal distribution?

```{r problem 2}
#Insert code for problem 2 here
```
# Include plot for problem 2 here

# Problem 3
# First create quintiles of age. Then draw a boxplot of BMI split by the age quintiles. Describe the association between BMI and age.

```{r problem 3}
#Insert code for problem 3 here
```
# Include plot for problem 3 here
# Interpret results for problem 3 here


