---
title: "EPID 674: Graphing in R"
subtitle: "Homework 3 Template"
author: "Insert name here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
format: html
editor: source
---


# Load packages

```{r}
#| label: load_packages
#| echo: true
#| results: 'hide'
#| message: false
#| warning: false

##### Load packages for this assignment, do this every time
library(tidyverse)
library(here)
library(ggpubr)
library(viridis)
library(RColorBrewer)
```

# Load data
```{r}
#| label: read_data

# Load saved NHANES homework dataset from the project files. This is the dataset we built in homework 2
load(here(("nhanes_homework_dataset.rda")), verbose = TRUE)
```


# 1. Create a scatterplot with a line of best fit

### Create a scatter plot with poverty-income ratio (variable: INDFMPIR) on the x-axis and digit symbol substitution test (variable: CFDDS) on the y-axis. Add a line of best fit for this relationship.

```{r}
#| label: scatterplot


# insert code here

```

Interpret the figure: 


# 2. Create a bar plot 

### Create a bar plot where cognitive impairment status (variable: MCI) is on the x-axis and the bar fill is by race/ethnicity (variable: race_eth). Try updating the bar colors. 

```{r}
#| label: barplot



# insert code here

```

Interpret the figure: 


# 3. Create a histogram or density plot

### Make a histogram or density plot with the digit symbol substitution test score (variable: CFDDS) on the x-axis. Have the fill and/or color of the plot by category of educational attainment.

```{r}
#| label: histogram_density_plot


# insert code here

```

Interpret the figure: Within each educational attainment group, the distribution of digit symbol substitution score is relatively normally distributed. 



# 4. Create a boxplot or violin plot

### Make a boxplot or violin plot with the digit symbol substitution test score (variable: CFDDS) on the y-axis. Split the boxes or violins by the age group (variable: age_group) on the x-axis. Try filling the boxes with the color of your choice

```{r}
#| label: boxplot



# insert code here

```

Interpret the figure: 



# 5. Create a multipaneled plot

### Create a multipaneled plot containing each of the previous plots from questions 1-4 (a barplot, a histogram, a boxplot or violin plot, and a scatterplot)

```{r}
#| label: multiplaneled

# insert code here

```

Interpret the figure: 
