---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Homework 5B"
author: "Insert Your Name Here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(foreign)
library(epiDisplay)
library(stats)
library(gmodels)
library(here)
```

# Using the dataset “bpa.rda”, answer the following. Include appropriate titles, labels and legends in all graphs.


# Problem 1
## Create a multi-frame plot and include 1) a bar chart of the race/ethnicity distribution; 2) a histogram of log-transformed HDL with males and females plotted separately; 3) boxplot of log-transformed urinary BPA by age quartiles; and 4) scatterplot with smoothing trend lines between age (in x-axis) and total cholesterol (in y-axis) by sex. 

```{r problem 1}
#Insert code for problem 1 here
```
# Include plot for problem 1 here

# Problem 2
# You want to describe the association between HDL (dependent variable, continuous) and urinary BPA (independent variable, continuous). First determine whether log-transformation for HDL is required. Create a scatterplot for HDL and urinary BPA. Add a least square line. What is the relationship between HDL and urinary BPA?

```{r problem 2}
#Insert code for problem 2 here
```
# Include plot for problem 2 here
# Interpret results for problem 2 here

# Problem 3
# Is the association between HDL and urinary BPA linear? Check this out using a smoothing plot (lowess) from the ggplot2 package or smooth.spline() in base graphics.

```{r problem 3}
#Insert code for problem 3 here
```
# Include plot for problem 3 here
# Interpret results for problem 3 here


