---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Graphing in R"
author: "Kelly Bakulski, Lauren Middleton"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
format: html
embed-resources: true
editor: source
---


# Install new packages
```{r}
#| label: install_packages
#| eval: false
#| include: false

# Install packages. Do this only once.

# To avoid installing every time you knit: keep chunk option #| eval: false

options(repos="https://cran.rstudio.com" )
install.packages("tidyverse")
install.packages("here")
install.packages("ggpubr")
install.packages("vidiris")
install.packages("RColorBrewer")

```


# Load packages
```{r}
#| label: load_packages
#| include: false

##### Load for the current session, do this every time
library(tidyverse)
library(here)
library(ggpubr)
library(viridis)
library(RColorBrewer)

```


# Load data
```{r}
#| label: read_data

# Check the file path
here("nhanes_class_dataset.rda")
# Load the saved R data
load(here("nhanes_class_dataset.rda"), verbose = TRUE)
```


# Useful websites for ggplot information:
* Online introduction book: **https://ggplot2-book.org/index.html**
* For theory: **https://rkabacoff.github.io/datavis/**
* How to code: **http://www.sthda.com/english/wiki/ggplot2-essentials*

# Introduction to ggplot: Scatter plot

```{r}
#| label: introduction_ggplot

# Specify the dataset, variables on the x-axis, y-axis
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI)) +
  geom_point()
```

# Customize a scatter plot

```{r}
#| label: customize_ggplot

# Update axes labels
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI)) +
  geom_point() + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL")

# Change point size, transparency, shape 
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI)) +
  geom_point(size = 0.3, 
             alpha = 0.2, 
             shape = 2) + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL")

# Add color variable
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           color = sex)) +
  geom_point(size = 0.4, 
             alpha = 0.2,
             shape = 2) + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL",
       color = "Sex")

# Change color palette
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           color = sex)) +
  geom_point(size = 0.4, 
             alpha = 0.2,
             shape = 2) + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL",
       color = "Sex of participant") +
  scale_color_viridis_d()

# Try another color palette
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           color = sex)) +
  geom_point(size = 0.4, 
             alpha = 0.2,
             shape = 2) + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL",
       color = "Sex") +
  scale_color_brewer(palette = "Spectral")

# Change theme to preset features
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           color = sex)) +
  geom_point(size = 0.4, 
             alpha = 0.2,
             shape = 2) + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL",
       color = "Sex") + 
  theme_classic()


# Manually change plot theme features
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           color = sex)) +
  geom_point(size = 0.4, 
             alpha = 0.2,
             shape = 2) + 
  labs(x = "Serum iron levels in ug/dL", 
       y = "Red blood cell count in million cells/uL",
       color = "Sex") + 
  theme(axis.text = element_text(color = "black", size = 10), #color and text size of axis labels
        axis.title = element_text(size = 12),   #text size of axis titles
        legend.title = element_text(size = 12), #text size of legend title
        legend.text = element_text(size = 10))  #text size of legend labels
```

# Check your understanding!
Make a scatter plot of cotinine (LBXCOT) on the x-axis and cadmium (LBXBCD) on the y-axis. Be sure to include axis labels with units. Experiment with changing the color, size, and transparency of the points. 


# Scatter plot with lines of best fit

```{r}
#| label: scatterplot_lines

# Add a linear fit
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI)) +
  geom_point() +
  geom_smooth(method = lm)

# Add a smoothing line
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI)) +
  geom_point() +
  geom_smooth()


# No points, just line and standard error
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI)) +
  geom_smooth(method = lm,
              linewidth = 1.5,
              se = TRUE,
              fill = "red") +
  labs(x = "Serum iron levels (ug/dL)",
       y = "Red blood cell count (million cells/uL)")

# Split by another variable
ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           shape = sex,
           color = sex)) +
  geom_point(size = 1,
             alpha = 0.3) +
  geom_smooth(method = lm,
              linewidth = 1.5) +
  labs(title = "Red blood cell count by iron and sex",
       x = "Serum iron levels (ug/dL)",
       y = "Red blood cell count (million cells/uL)")

```




# Make a barplot: Categorical data

```{r}
#| label: barplot

# Make a simple barplot
ggplot(nhanes,
       aes(x = sex)) +
  geom_bar()

# Add some colors
ggplot(nhanes,
       aes(x = sex,
           fill = sex,
           color = sex)) +
  geom_bar()

# Turn the plot horizontally
ggplot(nhanes,
       aes(x = sex,
           fill = sex,
           color = sex)) +
  geom_bar() +
  coord_flip()

# Add labels
ggplot(nhanes,
       aes(x = sex,
           fill = sex,
           color = sex)) +
  geom_bar() +
  labs(title = "Barplot by sex",
       x = "Sex",
       y = "Number of Participants")

# Manually set the colors
ggplot(nhanes,
       aes(x = sex,
           fill = sex)) +
  geom_bar() +
  labs(title = "Barplot by sex",
       x = "Sex",
       y = "Number of Participants") +
  scale_fill_manual(values=c("#999999", "#E69F00"))

# Plot two variables at once in stacked barchart
ggplot(nhanes,
       aes(x = sex,
           fill = education)) +
  geom_bar() +
  labs(title = "Barplot by sex and education",
       x = "Sex",
       y = "Number of Participants")

# Plot two variables at once in side by side barchart
ggplot(nhanes,
       aes(x = sex,
           fill = education)) +
  geom_bar(position = "dodge") +
  labs(title = "Barplot by sex and education",
       x = "Sex",
       y = "Number of Participants")

```



# Histogram: Numeric data

```{r}
#| label: histogram

# Histogram
ggplot(nhanes,
       aes(x = LBXRBCSI)) +
  geom_histogram()

# Adjust size of boxes
ggplot(nhanes,
       aes(x = LBXRBCSI)) +
  geom_histogram(binwidth = 0.25) + 
  labs("Red Blood Cell Count (million cells/uL)")

# Overlapping histograms
ggplot(nhanes,
       aes(x = LBXIRN,
           fill = sex)) + #stratify by sex
  geom_histogram(binwidth = 10,
                 position = "identity") #default position is stack

# Make a black and white plot with transparency
ggplot(nhanes,
       aes(x = LBXIRN,
           fill = sex)) +
  geom_histogram(binwidth = 10,
                 position = "identity",
                 alpha = 0.8) + #transparency
  scale_fill_grey(start = 0,
                  end = 0.75)
```


# Density plot

```{r}
#| label: density_plot

# Density plot
ggplot(nhanes,
       aes(x = LBXRBCSI)) +
  geom_density(size=2, 
               color="dodgerblue") +
  labs(x = "Red blood cell count (million cells/uL)")


# Add density plot to a histogram
ggplot(nhanes,
       aes(x = LBXRBCSI, y = ..density..)) +
  geom_histogram(fill = "lightgreen",
                 color = "seagreen",
                 binwidth = 0.25) +
  geom_density()

```


# Boxplot

```{r}
#| label: boxplot

# Boxplot
ggplot(nhanes,
       aes(x = LBXIRN)) +
  geom_boxplot(fill = "#990000",
               color = "#3366FF",
               notch = T,
               notchwidth = 0.5)

# Boxplot split by age groups
ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN)) +
  geom_boxplot(fill = "darkorchid") +
  labs(x = "Age groups",
       y = "Serum iron levels (ug/dL)")


# Boxplot split by age and sex, with colors!
ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN)) +
  geom_boxplot(aes(fill = sex)) +
  labs(title = "Boxplot of Iron Levels",
       x="Age",
       y="Iron levels")
```


# Violin plot

```{r}
#| label: violin_plot

# Basic violin plot
ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN)) + 
  geom_violin()

# Draw quantiles
ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN)) + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), 
              fill = "darkorchid",
              alpha = 0.2)

# Add beeswarm data points on top
ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN)) + 
  geom_violin(fill = "darkorchid", alpha = 0.2) +
  geom_jitter(shape = 16, #circle
              position = position_jitter(0.2),
              alpha = 0.15)

# Add colors
ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN,
           fill = age_groups)) + 
  geom_violin() +
  geom_jitter(shape = 16, #circle
              position = position_jitter(0.2),
              alpha = 0.15) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Violin plots by age group",
       x = "Age Group (years)",
       y = "Serum Iron Levels (ug/dL)")
```

# Check your understanding!

You want to visualize the relationship between iron status (iron_status) and red blood cells (LBXRBCSI). Determine an appropriate plot type, based on the variable types. Make a plot of these two variables. Practice updating the labels, colors, and other features.



# Facet wrap: Display categories in separate plots
 
```{r}
#| label: facet_wrap

# Stratified histograms in separate plots
ggplot(nhanes,
       aes(x = LBXIRN,
           fill = sex)) +
  geom_histogram() +
  facet_wrap(vars(sex),
             ncol = 1)

```


# Assign plots as objects

```{r}
#| label: assign_plot_objects

# Assign plots from earlier as objects
viol_plot <- ggplot(nhanes,
       aes(x = age_groups,
           y = LBXIRN,
           fill = age_groups)) + 
  geom_violin() +
  geom_jitter(shape = 16, #circle
              position = position_jitter(0.2),
              alpha = 0.25) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Violin plots by age group",
       x = "Age Group (years)",
       y = "Serum iron levels (ug/dL)")

scatter_plot <- ggplot(nhanes,
       aes(x = LBXIRN,
           y = LBXRBCSI,
           shape = sex)) +
  geom_point(aes(color = sex,
                 shape = sex),
             size = 1,
             alpha = 0.5) +
  geom_smooth(method = lm,
              aes(color = sex)) +
  labs(title = "Scatter plot red blood cells by iron and sex",
       x = "Serum iron levels",
       y = "Red blood cell count")

bar_plot <- ggplot(nhanes,
       aes(x = sex,
           fill = sex)) +
  geom_bar() +
  labs(title = "Barplot by sex",
       x = "Sex",
       y = "Number of Participants") +
  scale_fill_manual(values=c("#999999", "#E69F00"))

facet_plot <- ggplot(nhanes,
       aes(x = LBXIRN,
           fill = sex)) +
  geom_histogram() +
  facet_wrap(vars(sex),
             ncol = 1) + 
  labs(title = "Histogram of iron by sex",
         x = "Serum Iron in ug/dL")
```


# Combine multiple types of plots on one page

```{r}
#| label: multiplaneled_plots

# View the four plots on one page
ggarrange(bar_plot, viol_plot, scatter_plot, facet_plot,
          ncol = 2,
          nrow = 2,
          labels = LETTERS[1:4]) #add letters A-D as figure labels



# Assign compiled plot object
compiled_plots <- ggarrange(bar_plot, viol_plot, scatter_plot, facet_plot,
                            ncol = 2,
                            nrow = 2,
                            labels = LETTERS[1:4])
```


# Save and export plots

```{r}
#| label: saving_plots

# Recommended to save figures as both pdf for high-quality viewing and png for presentations

# Save the plot as a pdf for viewing at a high resolution
ggsave(filename = here("compiled_nhanes_plots.pdf"),
       plot = compiled_plots,
       width = 14,
       height = 9)

# Save the plot as a png for presentation with lower resolution
ggsave(filename = here("compiled_nhanes_plots.png"),
       plot = compiled_plots,
       units = "in",
       width = 14,
       height = 9,
       dpi = 300)
```


